Exploring the Vectorization of Python Constructs Using Pythran and Boost SIMD

### 2013-11-29 doc/papers/wpmvp14/paper.tex
Reading

A native module is a shared library whose interface make it callable from the
Python interpreter using the regular \texttt{import} mechanism. It makes it
possible to call native code from Python in a transparent way. Typically Python
object are unboxed when transfered from Python to C, where efficient
computations can happen. Then the result is boxed when getting back to the
Python world. In between, vectorization can happen. This is the approach taken
by the \texttt{numpy} module: a native type, the \texttt{ndarray} contains
contiguous unboxed elements, say single precision floating point numbers, and
provides a great deal of high-level functions to manipulate them. All the
computation-intensive part is done in pure C.  Vectorization can eventually
happen at this level, for instance when performing a point-to-point operation
like the sum of two arrays. However the granularity of the operations limits
the ratio of \texttt{load}/\texttt{store} versus operation. For instance, if
\texttt{a}, \texttt{b} and \texttt{c} are 1-D arrays, the sum \texttt{a + b +
c} is computed using two loops and an intermediate array. It means there are 4
load / stores and only two additions per loop iteration and vectorization is
unlikely to be profitable.

@@

Condition~\ref{enu:same-types} is a prerequisite for static typing and efficient
unboxing. For instance, any code that does not match it will fail to compile
using Pythran or Shed-Skin. It is always satisfied for the \texttt{ndarray}
container as it wraps unboxed arrays. Condition~\ref{enu:scalar-types} can be
checked at compile time. Python supports two native scalar types candidates for
vectorization: \texttt{int} (64 bits integer) and \texttt{float}
(double-precision floating point numbers). The \texttt{numpy} module adds support for a
wide range of signed and unsigned integers as well as single, double or
quadruple precision float. Depending on the target architecture, only some of
them can fit into vector registers. Finally condition~\ref{enu:contiguous} also
is a static property of a container. It is verified for C++ alternative to
Python's \texttt{list}, \texttt{str} and \texttt{ndarrays}, but generally not
for \texttt{set} or \texttt{dict}. Condition \ref{enu:random-access} is not
matched by \emph{generator}s, the Python incarnation of co-routines.

%we know a work have to be done about random acces but you speak about something else now ...
%It is disturbing and we don't really know why you speak about this new topic.

@@

The typical way to solve this issue in C++ is to use \emph{expression
templates}, a technique described in~\cite{expression_templates} that uses the
C++ type system to perform lazy evaluation of an expression. Yet, this
approach is limited to a single expression, as illustrated by Listing~\ref{lst:numpy-code-fs}: in that case expression templates would not prevent the creation
of temporary arrays \texttt{t0} and \texttt{t1}. Fortunately, it is possible to
apply \emph{forward substitution} to fall-back to the expression from
Listing~\ref{lst:numpy-code} and avoid the creation of any temporary in some
case detected with static analysis of the source code.

>>>>

A native module is a shared library whose interface make it callable from the
Python interpreter using the regular \texttt{import} mechanism. It makes it
possible to call native code from Python in a transparent way. Typically Python
object are unboxed when transfered from Python to C, where efficient
computations can happen. Then the result is boxed when getting back to the
Python world. In between, vectorization can happen. This is the approach taken
by the \texttt{numpy} module: a native type, the \texttt{ndarray} contains
contiguous unboxed elements, say single precision floating point numbers, and
provides a great deal of high-level functions to manipulate them. All the
computation-intensive part is done in pure C.  Vectorization can eventually
happen at this level, for instance when performing a point-to-point operation
like the sum of two arrays. However the granularity of the operations limits
the ratio of \texttt{load}/\texttt{store} versus operation. For instance, if
\texttt{a}, \texttt{b} and \texttt{c} are 1-D arrays, the sum \texttt{a + b +
c} is computed using two loops and an intermediate array. It means there are 4
\texttt{load}/\texttt{store} and only two additions per loop iteration and vectorization is
unlikely to be profitable.

@@

Condition~\ref{enu:same-types} is a prerequisite for static typing and efficient
unboxing. For instance, any code that does not match it will fail to compile
using Pythran or Shed-Skin. It is always satisfied for the \texttt{ndarray}
container as it wraps unboxed arrays. Condition~\ref{enu:scalar-types} can be
checked at compile time. Python supports two native scalar types candidates for
vectorization: \texttt{int} (64 bits integer) and \texttt{float}
(double-precision floating point numbers). The \texttt{numpy} module adds support for a
wide range of signed and unsigned integers as well as single, double or
quadruple precision float. Depending on the target architecture, only some of
them can fit into vector registers. Finally condition~\ref{enu:contiguous} also
is a static property of a container. It is verified for C++ alternative to
Python's \texttt{list}, \texttt{str} and \texttt{ndarrays}, but generally not
for \texttt{set} or \texttt{dict}. Condition \ref{enu:random-access} is matched by
the C++ alternative of contiguous container but it is not
matched by \emph{generator}s, the Python incarnation of co-routines.

@@

The typical way to solve this issue in C++ is to use \emph{expression
templates}, a technique described in~\cite{expression_templates} that uses the
C++ type system to perform lazy evaluation of an expression. Yet, this
approach is limited to a single expression, as illustrated by Listing~\ref{lst:numpy-code-fs}: in that case expression templates would not prevent the creation
of temporary arrays \texttt{t0} and \texttt{t1}. Fortunately, it is possible to
apply \emph{forward substitution} to fall-back to the expression from
Listing~\ref{lst:numpy-code} and avoid the creation of any temporary based on
static analysis of the source code.
